---
title: "R Notebook for Assessing the Physiological Consequences of Current Management Practices on African Elephants (Loxodonta africana): A Case study from the Nama-Karoo"
author: "Kyla Funk"
date: "September 12 2024"

---
###  Body Condition of Translocated African Elephants (Loxodonta africana) in a Semi-Arid Environment: Effects of Seasonality, Age-Sex, and Management Interventions  
#### Kyla R. Funk 1, Leslie R. Brown 2, Eric Vander Wal 3
# 1 Department of Cognitive and Behavioural Ecology, Memorial University of Newfoundland, Canada
# 2 Applied Behavioural Ecology and Ecosystems Research Unit, University of South Africa, Pretoria, South Africa
# 3 Department of Biology, Memorial University of Newfoundland, Canada


#### Correspondence concerning this article should be addressed to Kyla Funk, Address: Memorial University of Newfoundland, Department of Cognitive and Behavioural Ecology, 1 Arctic Ave, St. John’s, NL, Canada. E-mail: krfunk@mun.ca. Phone: +14037973865

---
##Load libraries
```{r}
library(netTS)
library(lubridate)
library(ggplot2)
library(reshape2)
library(readxl)
library(dplyr)
library(bayesplot)
library(see)
library(bayestestR)
library(brms)
library(DHARMa)
library(performance)
library(stringr)
library(ggridges)
library(RColorBrewer)
library(ggpubr)
library(tidybayes)
library(modelr)
```

##look at data
```{r load data and plot body condition with other variables}
BCMonth <- read_excel("FUNK_2024_DataBook.xlsx", sheet = "RawData")

BCMonth <- BCMonth %>% select(
  Score_Average,
  D_W,
  Dry.Wet_Season,
  AgeSex,
  Below_Above_Average,
  ID,
  Rainfall_Season,
  Difference_in_Average,
  ObservationTotalSeason
)

BCMonth <- transform(BCMonth, Score_Average = as.numeric(Score_Average))
BCMonth <- transform(BCMonth, D_W = as.factor(D_W))
BCMonth <- transform(BCMonth, Dry.Wet_Season = as.factor(Dry.Wet_Season))
BCMonth <- transform(BCMonth, AgeSex = as.factor(AgeSex))
BCMonth <- transform(BCMonth, Below_Above_Average = as.factor(Below_Above_Average))

BCMonth$Rainfall.s <- scale(BCMonth$Rainfall_Season, center = TRUE, scale = TRUE)
BCMonth$Average.Diff.s <- scale(BCMonth$Difference_in_Average,
                                center = TRUE,
                                scale = TRUE)


BCMonth <- na.omit(BCMonth)

ggplot(BCMonth, aes(x = AgeSex, y = Score_Average)) +  geom_boxplot(fill =
                                                                      'blue')
ggplot(BCMonth, aes(x = Age, y = Score_Average)) +  geom_boxplot(fill =
                                                                   'blue')
ggplot(BCMonth, aes(x = Sex, y = Score_Average)) +  geom_boxplot(fill =
                                                                   'blue')
ggplot(BCMonth, aes(x = D_W, y = Score_Average)) +  geom_boxplot(fill =
                                                                   'blue')
ggplot(BCMonth, aes(x = Dry.Wet_Season, y = Score_Average)) +  geom_boxplot(fill =
                                                                              'blue')
ggplot(BCMonth, aes(x = Below_Above_Average, y = Score_Average)) +  geom_boxplot(fill =
                                                                                   'blue')
```


##Build beta model, is body score average predicted by age-sex category, pooled dry seasons vs pooled wet seaosns, individual season, rainfall, if the rainfall was above or below avergae, or difference in rainfall from the average rainfall? 
```{r agesex model}

priors.BRMS.interceptb.norm<- c(
 
  set_prior("normal(0, 1)", class = "b")
)

M2.BC <- bf(Score_Average ~ AgeSex + D_W + Dry.Wet_Season + Rainfall.s + Below_Above_Average + Average.Diff.s + (1|ID) + offset(log(ObservationTotalSeason))) 
          
fit.M2.BC <- brm(M2.BC, 
                  data = BCMonth, 
                      family = Beta,
                            prior=priors.BRMS.interceptb.norm, 
                                  chains = 4,iter = 2000, 
                                      control=list(
                                            max_treedepth = 15), cores = 4)
```

##checking the Beta model: is it a good fit?
```{r dispersion checks and summary.}
# save(fit.M1.2.BC, file = "BodyConditonModel.Rdata")
# fit.M2.BC <- load("~/Documents/R_Projects/Elephant_BC_Data/Elephant_BC/BodyConditonModel.Rdata")
bayesplot::color_scheme_set("teal")


#PP_Checks for SM
# Figure S1
#tiff("M2.PP_Check.tiff", units="in", width=4, height=3, res=400)
pp_check(fit.M2.BC) + theme(
  panel.background = element_rect(fill = "transparent", colour = NA),
  axis.line = element_line(
    size = 0.5,
    colour = "black",
    linetype = 1
  )
) +  scale_fill_manual(values = c(pal2[2:3]), name = NULL)

#tiff("M2.PP_Check2.tiff", units="in", width=4, height=3, res=400)
pp_check(fit.M2.BC, type = "dens") + theme(panel.background = element_rect(fill = "transparent", colour = NA))

#Dispersion Tests for SM
# Figure S2 A
#tiff("dispersion.tiff", units="in", width=7, height=5, res=400)
PosteriorPredictions.fit.M2.BC <- predict(fit.M2.BC, summary = FALSE)
PosteriorPredictions.fit.M2.BC <- t(PosteriorPredictions.fit.M2.BC)
medianPosteriorPredictions.fit.M2.BC <- apply(PosteriorPredictions.fit.M2.BC, 1, median)
observations.fit.M2.BC <- BCMonth$Score_Average
DharmaM1.bayes.fit.M2.BC <- createDHARMa(
  fittedPredictedResponse = medianPosteriorPredictions.fit.M2.BC,
  simulatedResponse = PosteriorPredictions.fit.M2.BC,
  observedResponse = observations.fit.M2.BC,
  integerResponse = F
)
testDispersion.fit.M2.BC <- testDispersion(DharmaM1.bayes.fit.M2.BC, plot =
                                             T)
# Figure S2 B
#tiff("dispersion2.tiff", units="in", width=7, height=5, res=400)
plotResiduals(DharmaM1.bayes.fit.M2.BC)

#Summary of output
# Table 3
summary(fit.M2.BC)
#Probability of Direction for BC model
p_direction(fit.M2.BC, method = "direct", null = 0)
```
#Bayesplot 
```{r Bayesplot - visual summary of Table 3}
#tiff("M1.Poisson.estimates.tiff", units="in", width=4, height=3, res=400)
bayesplot::mcmc_areas(
  as.matrix(fit.M2.BC),
  regex_pars = c(
    "b_Intercept",
    "b_AgeSexFJ",
    "b_AgeSexMA",
    "b_AgeSexMJ",
    "b_D_WW",
    "b_Dry.Wet_Seasonb.W1",
    "b_Dry.Wet_Seasonc.D2",
    "b_Dry.Wet_Seasond.W2",
    "b_Dry.Wet_Seasone.D3",
    "b_Dry.Wet_Seasonf.W3",
    "b_Dry.Wet_Seasong.D4",
    "b_Dry.Wet_Seasonh.W4",
    "b_Rainfall.s",
    "b_Below_Above_AverageB",
    "b_Average.Diff.s"
  ),
  prob = 0.95
) + geom_vline(xintercept = 0, linetype = "dashed") + theme(
  axis.ticks.length = unit(-0.25, "cm"),
  panel.border = element_rect(
    colour = "black",
    fill = NA,
    size = 1
  ),
  axis.line = element_line(colour = 'black', size = 1),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  axis.text = element_text(
    size = 5,
    color = "black",
    family = "Helvetica"
  ),
  panel.background = element_rect(fill = "transparent", colour = NA),
  plot.background = element_rect(fill = "transparent", colour = NA),
  axis.title.x = element_text(family = "Helvetica", size = 10),
  legend.position = "none"
) + xlim(-4, 3)
```
#getting R2 for Beta model
```{r conditional and marginal r2}
m.2.conditional.fit.M2.BC <- bayes_R2(fit.M2.BC)
m.2.marginal.fit.M2.BC <- bayes_R2(fit.M2.BC, re_formula = NA)
m.2.conditional.fit.M2.BC
m.2.marginal.fit.M2.BC
```

#Plot posterior expectations
```{r}
# Expectation draws
BC_epred <- fit.M2.BC |>
  add_epred_draws(newdata = BCMonth)

# Figure 1. Plot posterior expectations for relationship between agesex and body score average
# tiff("PosteriorExpectationsAgeSex.tiff", units="in", width=7, height=5, res=400)
ggplot(BC_epred) +
  stat_halfeye(aes(AgeSex, .epred), alpha = 0.5) +
  labs(x = 'Age-Sex Category', y = 'Body Condition Score Average') +
  theme(
    panel.background = element_rect(fill = "transparent", colour = NA),
    axis.line = element_line(
      size = 0.5,
      colour = "black",
      linetype = 0.5
    ),
    axis.ticks.length = unit(-0.25, "cm"),
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      size = 1
    )
  ) +
  scale_x_discrete(
    labels = c(
      "FA" = "Adult females",
      "FJ" = "Juvenile females",
      "MA" = "Adult males",
      "MJ" = "Juvenile males"
    )
  )

# Figure 2. Plot posterior expectations for relationship between Season and body score average
#tiff("PosteriorExpectationsSeason.tiff", units="in", width=7, height=5, res=400)
ggplot(BC_epred) +
  stat_halfeye(aes(Dry.Wet_Season, .epred), alpha = 0.5) +
  labs(x = 'Season', y = 'Body Score Average') +
  theme(
    panel.background = element_rect(fill = "transparent", colour = NA),
    axis.line = element_line(
      size = 0.5,
      colour = "black",
      linetype = 0.5
    ),
    axis.ticks.length = unit(-0.25, "cm"),
    panel.border = element_rect(
      colour = "black",
      fill = NA,
      size = 1
    )
  ) +
  scale_x_discrete(
    labels = c(
      "a.D1" = "DS 1",
      "b.W1" = "WS 1",
      "c.D2" = "DS 2",
      "d.W2" = "WS 2",
      "e.D3" = "DS 3",
      "f.W3" = "WS 3",
      "g.D4" = "DS 4",
      "h.W4" = "WS 4"
    )
  )

# Plot posterior expectations for relationship between above or below average rainfall and body score average
ggplot(BC_epred) +
  stat_halfeye(aes(Below_Above_Average, .epred), alpha = 0.5) +
  labs(x = 'Below or Above average rainfall', y = 'Body Score Average')

```
